Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-07-29 manuscripts/reply/reply-main.tex
Working on overhaul of intro for reply ms.

\section{Introduction}
Biogeographers frequently seek to explain population and species
differentiation on geographical phenomena.
Establishing that a set of population splitting events occurred
at the same time can be a potentially persuasive argument that a set of taxa
were affected by the same geographic events.
% Unfortunately, estimating divergence times precisely when one lacks
% precise information about the rate of molecular evolution is a notoriously
% difficult problem in evolutionary biology.
% \citet{Huang2011} introduced an approximate Bayesian approach (implemented in
% the software \msb) to estimating the probabilities of models in which multiple
% sets of taxa diverge at the same time. 
The approximate-Bayesian method, \msb, allows biogeographers to estimate the
probabilities of models in which multiple sets of taxa diverge at the same
time \citep{Hickerson2006,Huang2011}.
% This method  does not require a common set of character data that links all
% of the taxa under investigation, but does require the relative mutation
% rates among the taxa and loci to be assumed \emph{a priori}.

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method was very sensitive to prior assumptions and suggested
this was due to the fact that the marginal likelihoods of divergence models are
sensitive to the use of uniform priors on divergence times.
They also suggested alternative probability distributions (e.g., gamma) could
better represent prior uncertainty about divergence times while mitigating the
disparity of marginal likelihoods across divergence models.
In response, \citet{Hickerson2013} suggested the lack of robustness in \msb
analyses was caused by sampling error due to the inefficiency of the method's
rejection algorithm.
As a potential solution, they proposed a model-averaging approach that uses
narrow, empirically informed uniform priors.

Whereas the papers of \citet{Oaks2012} and \citet{Hickerson2013} differ about
the cause and solution, they agree on the fundamental methodological point
about the model selection performed in \msb:
% The papers of \citet{Oaks2012} and \citet{Hickerson2013} are in agreement
% on the fundamental methodological point about the model selection performed in \msb:
\begin{itemize}
   \item The use of vague priors on the divergence-time parameters can lead to
       support for models with few divergence events shared across taxa. Thus,
       the primary inference enabled by the approach is very sensitive to the
       priors on divergence times.
\end{itemize}
This is not surprising given the rich statistical literature showing that
marginal likelihoods are very sensitive to the priors used in Bayesian model
selection
\citep[e.g.,][]{Jeffreys1939,Lindley1957}.
% We know that under broad uniform priors on divergence times, models with few
% divergence parameters will have greater marginal likelihoods relative to more
% parameter-rich models.
% Accordingly, \citet{Oaks2012} suggest the primary cause of spurious support
% for models with few divergence parameters is the greater marginal likelihoods
% of these models under vague uniform priors, relative to more parameter-rich
% models.
Under vague uniform priors on divergence times, models with more
divergence-time parameters integrate over \emph{much} greater parameter space
with low probability of producing the data, yet relatively high prior density
imposed by the uniformly distributed prior.
% By using such priors on divergence times, we are essentially telling the
% method to favor models with few divergence times even if they do not
% explain the data very well.
Thus, uniform priors on divergence times can impose a strong ``penalty'' for
divergence-time parameters, creating a strong preference for simple
models that are often of interest to biogeographers, even if these models do
not explain the data very well.
% It is important to note that this is not due to estimation error
% (i.e., this phenomenon is extrinsic to the numerical approximation machinery of
% \msb).
% Rather, the \emph{true} posterior will tend to support models with few
% divergence-time parameters under broad uniform priors on divergence times.
Such posterior support for simultaneous divergence, even if correct from the
perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.
% Ideally, the estimated posterior probability of a simultaneous-divergence model
% can be interpreted as an estimate of the probability that this model is correct.
% However, when using uniform priors on divergence times, it might be better
% interpreted as an estimate of the probability that the simultaneous-divergence
% model cannot be completely ruled out as a possible explanation of the data.
% Note, this is a statistical issue, extrinsic to \msb, and does not question the
% soundness of the numerical approximation machinery of \msb.

% Alternatively, \citet{Hickerson2013} suggest the problem is the result of a
% numerical issue intrinsic to \msb, concluding the method's rejection algorithm
% is inefficient and will be biased toward models with fewer divergence-time
% parameters, because the parameter space of these models are more densely
% sampled relative to models with more divergence-time parameters.
% We agree that sampling error is present in all numerical Bayesian estimation
% methods, however, we show here that it is not a major contributer to the
% biases found by \citet{Oaks2012}.
% We present results here of analyses that, when combined with the results
% of \citet{Oaks2012}, strongly suggest that \msb prefers models with few
% divergence-time parameters, because they have greater likelihoods when
% averaged over broad uniform priors on divergences times.

>>>>

\section{Introduction}
Biogeographers frequently seek to explain population and species
differentiation on geographical phenomena.
Establishing that a set of population splitting events occurred
at the same time can be a potentially persuasive argument that a set of taxa
were affected by the same geographic events.
The approximate-Bayesian method, \msb, allows biogeographers to estimate the
probabilities of models in which multiple sets of taxa diverge at the same
time \citep{Hickerson2006,Huang2011}.

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method was very sensitive to prior assumptions and often
incorrectly supported shared divergences across taxa, to which
\citet{Hickerson2013} responded.
Both papers agree on the fundamental methodological point about the model
selection performed in \msb:
% The papers of \citet{Oaks2012} and \citet{Hickerson2013} are in agreement
% on the fundamental methodological point about the model selection performed in \msb:
\begin{itemize}
   \item The use of vague priors on the divergence-time parameters can lead to
       support for models with few divergence events shared across taxa. Thus,
       the primary inference enabled by the approach is very sensitive to the
       priors on divergence times.
\end{itemize}
However, the two papers suggest alternative causes for this behavior:
\begin{itemize}
    \item \citep{Hickerson2013} suggest broad uniform priors cause numerical
        problems in which the ABC rejection algorithm is unable to sufficiently
        sample the space of the models in reasonable computational time. The
        method will incorrectly support models with fewer divergence-time
        parameters because they are better sampled.
    \item \citet{Oaks2012} suggest the cause is more fundamental; the broad
        uniform priors simply lead to very small marginal likelihoods of models
        with many divergence-time parameters.
\end{itemize}
In the former case, the true posterior would support heterogeneity in 
divergences among taxa in the face of temporally random divergences; the
problem only arises from our inability to accurately estimate the posterior.
In the latter case, the true posterior would support simultaneous divergences
among taxa in the face of random divergences. I.e., under the prior
assumptions about divergence times, this is the correct answer from 
the perspective of Bayesian model choice.
Such posterior support for simultaneous divergence, even if correct, does not
provide the biogeographical insights that a researcher who employs \msb seeks
to gain.

The distinction between these potential causes is important for improving our
ability to estimate shared divergence histories.
If the cause is numerical, then the model is sound and we simply need to
improve our estimation procedures.
For example, \citet{Hickerson2013} proposed a model-averaging approach that
uses narrow, empirically informed uniform priors in the hope that with less
parameter space to sample, the rejection algorithm will produce better
estimates of the posterior.
If the cause is marginal likelihoods, we need to correct the model, because
improving the estimation machinery will only increase our ability to estimate
biogeographically misleading posteriors.
Accordingly, \citet{Oaks2014dpp} has introduced a method that uses more
flexible probability distributions to accommodate prior uncertainty in
divergence times without overly inhibiting the marginal likelihoods of models
with more divergence-time parameters, which increases the method's robustness
and power to detect temporal variation in divergences.

This is not surprising given the rich statistical literature showing that
marginal likelihoods are very sensitive to the priors used in Bayesian model
selection \citep[e.g.,][]{Jeffreys1939,Lindley1957}.
Thus, uniform priors on divergence times can impose a strong ``penalty'' for
divergence-time parameters, creating a strong preference for simple
models that are often of interest to biogeographers, even if these models do
not explain the data very well.
